Generating gaits for physical quadruped robots: evolved neural networks vs. local parameterized search

### 2011-04-14 paper_ecal/experimentalSetup.tex
more forging

An important aspect of this study was to evaluate gait generation and
learning algorithms in the noisy, unforgiving real world instead of
using a simulator.  In this section we give details of the hardware
setup we used.  First we discuss the robotic platform on which gaits
were run, and then we describe the tracking system used to quantify
the fitness of those gaits.

>>>>

We next describe the details of the experimental system.

### 2011-04-14 paper_ecal/fitnessEvaluation.tex
more forging

\acmFigg{wiiMote_crop.jpg}{robot_led_crop.jpg}{1}{A Nintendo Wii
  remote provided the location of the robot by tracking the infrared
  LED mounted on the robot's antenna.  The LED position was measured
  in pixels and transmitted from the Wii remote to the robot via
  bluetooth.}

@@

A separate tracking server, written in Python, ran on the robot and
interfaced with the Wii remote via bluetooth using the \code{CWiid}
library \citep{cwiid}.  Our fitness testing code communicated with this
server via a socket connection and requested position updates at the
beginning and end of each run.

As mentioned earlier, the metric for evaluating gaits was their
average speed, which was measured as the Euclidian distance the robot
moved during a 12-second run. For the manual and parameterized gaits,
the fitness was this value. The HyperNEAT gaits stressed the motors
more than the other gaits, so to encourage gaits that did not tax the
motors we penalized gaits that caused the servos to stop
responding. When the servos stopped responding they could, in nearly
all cases, be restarted by cycling power, though over the course of
this study we did have to replace four servos that were damaged.
The penalty was to set the fitness to half of the distance
the robot actually traveled. We tested whether the servos were
responding after each gait by commanding them to specific positions
and checking whether they actually moved to those positions. This test
had the additional benefit of rewarding those gaits that did not flip
the robot into a position where it could not move its legs, which
HyperNEAT also did more than the other learning methods. Because the
fitness of HyperNEAT gaits were often halved, in results we compare
actual distance traveled in addition to fitness for the best gaits
produced by each class of gait-generating algorithms.

@@

The only human intervention required during most learning trials was
to occasionally move the robot back into the viewable area of the Wii
remote whenever it wandered outside this window.  Initially this was a rare
occurrence, as the gaits did not typically produce motion as large as
the size of the window (roughly 175 x 120 cm).  However, as gaits
improved, particularly when using HyperNEAT, the robot began to walk
out of the measurement area a non-negligible fraction of the time.
Whenever it did so, we would discard the trial and repeat it until the
gait was entirely within the window.

While this process guaranteed that we always obtained a measurement
for a given gait before proceeding, it also biased the measurements
downward.  Because the performance of the robot on a given gait varied
from trial to trial, a successful measurement was more likely to be
obtained when the gait happened to perform poorly.  This phenomenon
was negligible at first, but became significantly more pronounced when
evaluating several gaits that produced motion across nearly the entire
area.  HyperNEAT gaits were especially likely to require additional trials, meaning that the reported performance for HyperNEAT is worse than it would have been otherwise. For future studies we will likely use a 2x2 array of Wii
remotes to double the length and width of the measurement arena.

>>>>

\acmFigg{wiiMote_crop.jpg}{robot_led_crop.jpg}{1}{A Nintendo Wii
  remote provided the location of the robot by tracking the infrared
  LED mounted on the robot's antenna.  The position was measured
  in pixels and transmitted from the Wii remote to the robot via
  bluetooth.}

@@

A separate Python tracking server ran on the robot and
interfaced with the Wii remote via bluetooth using the \code{CWiid}
library \citep{cwiid}.  Our fitness-testing code communicated with this
server via a socket connection and requested position updates at the
beginning and end of each run.

As mentioned earlier, the metric for evaluating gaits was the Euclidian distance the robot
moved during a 12-second run. For the manual and parameterized gaits,
the fitness was this value. The HyperNEAT gaits stressed the motors
more than the other gaits, so to encourage gaits that did not tax the
motors we penalized gaits that caused the servos to stop
responding. When the servos stopped responding they could, in nearly
all cases, be restarted by cycling power, though over the course of
this study we did have to replace four servos that were damaged.
The penalty was to set the fitness to half of the distance
the robot actually traveled. We tested whether the servos were
responding after each gait by commanding them to specific positions
and checking whether they actually moved to those positions. This test
had the additional benefit of rewarding those gaits that did not flip
the robot into a position where it could not move its legs, which
HyperNEAT also did more than the other learning methods. Because the
fitness of HyperNEAT gaits were often halved, in results we compare
actual distance traveled in addition to fitness for the best gaits
produced by each class of gait-generating algorithms.

@@

The only human intervention required during most learning trials was
to occasionally move the robot back into the viewable area of the Wii
remote whenever it left this window.  Initially this was a rare
occurrence, as the gaits did not typically produce motion as large as
the size of the window (roughly 175 x 120 cm).  However, as gaits
improved, particularly when using HyperNEAT, the robot began to walk
out of the measurement area a non-negligible fraction of the time.
Whenever it did so, we would discard the trial and repeat it until the
gait finished within the window. While this process guaranteed that we always obtained a measurement
for a given gait before proceeding, it also biased some measurements
downward.  Because the performance of the robot on a given gait varied
from trial to trial, a successful measurement was more likely to be
obtained when the gait happened to perform poorly.  This phenomenon
was negligible at first, but became more pronounced as gaits began traversing the entire
area.  HyperNEAT gaits were especially likely to require additional trials, meaning that the reported performance for HyperNEAT is worse than it would have been otherwise. For future studies we will use a $2\times2$ array of Wii
remotes to double the size of the measurement arena.

### 2011-04-14 paper_ecal/introduction.tex
more forging

An additional motivation is to test whether gaits evolved in simulation transfer to reality well, especially those evolved with cutting-edge evolutionary algorithms. Because HyperNEAT gaits performed well in
simulation, it is interesting to test whether HyperNEAT can produce fast
gaits for a physical robot. Such tests help us better understand the real world implications of results reported only in simulation. It is additionally interesting to test how
more traditional gait optimization techniques compete with
evolutionary algorithms when evolving in hardware. A final motivation of this research is simply to evolve effective gaits for a physical robot.

>>>>

An additional motivation is to test whether gaits evolved in simulation transfer to reality well, especially those evolved with cutting-edge evolutionary algorithms. Because HyperNEAT gaits performed well in
simulation, it is interesting to test whether HyperNEAT can produce fast
gaits for a physical robot, including handling the noisy, unforgiving nature of the real world. Such tests help us better understand the real world implications of results reported only in simulation. It is additionally interesting to test how
more traditional gait optimization techniques compete with
evolutionary algorithms when evolving in hardware. A final motivation of this research is simply to evolve effective gaits for a physical robot.

### 2011-04-14 paper_ecal/learningMethods.tex
More edits from reviews, made figures larger
more forging

We have chosen to implement two classes of algorithms to produce
the motion-generating function $g(t)$ (described earlier).  The
first method uses a family of parameterized gaits, described in this
section and the next. The second method, HyperNEAT, evolves a non-parametric gait
controlled by neural networks and is described last.

@@

By a \emph{parameterized gait}, we mean a gait produced by a
parameterized function $g(t; \vec{\theta})$. Fixing the parameter
$\vec{\theta}$ yields a deterministic motion function over time.  We
tried several parametrizations on the robot and, upon obtaining
reasonable early success, settled on one particular parametrization, which we call  \emph{SineModel5}. Its root pattern is a sine wave and it has five parameters (\tabref{params}).

@@

\begin{table}[h!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameter    & Description               & Range \\
\hline
\hline
\amp         & Amplitude                 & [0, 400] \\
\hline
$\tau$       & Period                    & [.5, 8] \\
\hline
$m_O$        & Outer-motor multiplier    & [-2, 2] \\
\hline
$m_F$        & Front-motor multiplier    & [-1, 1] \\
\hline
$m_R$        & Right-motor multiplier    & [-1, 1] \\
\hline
\end{tabular}
\caption{The \emph{SineModel5} motion model parameters.}
\tablabel{parameters}
\label{tab:params}
\end{center}
\end{table}

Intuitively, SineModel5 starts with 8 identical sine waves of
amplitude $\amp$ and period $\tau$, multiplies the waves for all outer
motors by $m_O$, multiplies the waves for all front motors by $m_F$,
and multiplies the waves for all right motors by $m_R$.  To obtain the
actual motor position commands, these waves are offset by
appropriate fixed constants so that the base position (when the sine
waves are at 0) is approximately a crouch (the position shown in
\figref{robot_whitebg.jpg}).  In this motion model, the ninth
(center) motor is kept at a neutral position.  Thus, the commanded
position for all motors, as a vector function of time, is:

@@

\noindent Here the fixed constants that determine the crouched
position are $C_I = 800$ for inner motors, $C_O = 40$ for outer
motors, and $C_C = 512$ for the center hip motor.  The nine functions (in order)
correspond to the nine motors listed in
\figref{topdown2.jpg}.

@@

Using the SineModel5 parameterized motion model described in the
previous section, along with the allowable ranges for each of the five
parameters (shown in \tabref{parameters}), the task becomes how to
choose the combination of five parameters that results in the fastest
motion, per the chosen evaluation methods described earlier.

@@

We evaluated a total of six different
learning algorithms for the parameterized motion models.  All methods
needed a way of picking a $\vec{\theta}$ for the first trial of each run.  This could be done by selecting a random starting vector within the
allowable domain.  However, to most directly compare learning methods
by their ability to improve on the initially chosen $\vec{\theta}$, we
decided to evaluate the different methods by starting them at the same three
initial $\vec{\theta}$ vectors in the three runs.

@@

The six learning algorithms for the parameterized motion models are as follows:

@@

\emph{Random}: This method randomly generates parameter vectors in the allowable range for every trial.  This strategy was used only
  as baseline for comparison.

\emph{Uniform random hill climbing}: This method always starts
  with the current best gait, and then chooses next $\vec{\theta}$ by
  randomly choosing one parameter to adjust and replacing it with a
  new value chosen with uniform probability in the allowable range for
  that parameter. This new point is evaluated, and if it results in a
  longer distance walked than the previous best gait, it is saved as
  the new best gait. The process is then repeated, always starting
  with the best gait.

@@

\emph{N-dimensional policy gradient ascent}: We implemented Kohl
  and Stone's \citep{kohl} method for local gradient ascent for gait
  learning with noisy fitness evaluations. This strategy explicitly
  estimates the gradient of the objective function. It does this by
  first generating $n$ parameter vectors near the initial vector by
  perturbing each dimension of each vector randomly by either
  $-\epsilon$, $0$, or $\epsilon$. Then each vector is run on the
  robot, and for each dimension we group the results into three
  groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along this
  dimension is then estimated as the average score for the $\epsilon$
  group minus the average score for the $-\epsilon$ group. Finally,
  the method creates the next $\vec{\theta}$ by changing all
  parameters by a fixed-size step in the direction of the gradient.
  For this study we used values of $\epsilon$ equal to 5\% of the
  allowable range in each dimension (ranges listed in
  \tabref{parameters}), and a step size scaled such that if all dimensions
  were in the range [0,~1], the norm of the step size would be 0.1.

@@

Each of the six learning methods was run on 3 different initial
parameter vectors ($\vec{\theta}_A$,
$\vec{\theta}_B$, and $\vec{\theta}_C$) in order to fairly compare the algorithms. Three runs were performed per learning method. Runs continued until the performance plateaued, which we defined as when there was no improvement during the last third of
a run.

>>>>

We implemented two classes of algorithms to produce
the motion-generating function $g(t)$ (described earlier).  The
first method uses a family of parameterized gaits, described in this
section and the next. The second method, HyperNEAT, evolves a non-parametric gait
controlled by neural networks and is described last.

@@

By a \emph{parameterized gait}, we mean a gait produced by a
parameterized function $g(t; \vec{\theta})$. Fixing the parameters
$\vec{\theta}$ yields a deterministic motion function over time.  We
tried several parametrizations on the robot and, upon obtaining
reasonable early success, settled on one particular parametrization, which we call  \emph{SineModel5}. Its root pattern is a sine wave and it has five parameters (\tabref{params}).

@@

\begin{table}[h!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameters        &                           &       \\
in $\vec{\theta}$ & Description               & Range \\
\hline
\hline
\amp         & Amplitude                 & [0, 400] \\
\hline
$\tau$       & Period                    & [.5, 8] \\
\hline
$m_O$        & Outer-motor multiplier    & [-2, 2] \\
\hline
$m_F$        & Front-motor multiplier    & [-1, 1] \\
\hline
$m_R$        & Right-motor multiplier    & [-1, 1] \\
\hline
\end{tabular}
\caption{The \emph{SineModel5} motion model parameters.}
\tablabel{parameters}
\label{tab:params}
\end{center}
\end{table}

Intuitively, SineModel5 starts with 8 identical sine waves of
amplitude $\amp$ and period $\tau$, multiplies the waves for all outer
motors by $m_O$, multiplies the waves for all front motors by $m_F$,
and multiplies the waves for all right motors by $m_R$.  To obtain the
actual motor position commands, these waves are offset by
appropriate fixed constants ($C_O = 40$ for outer
motors, $C_I = 800$ for inner motors, and $C_C = 512$ for the center hip motor) so that the base position (when the sine
waves are at 0) is approximately a crouch (the position shown in
\figref{robot_whitebg.jpg}).  In this motion model, the ninth
(center) motor is kept at a neutral position.  Thus, the commanded
position for each motor (numbered in \figref{topdown2.jpg}), as a vector function of time, are produced by the following nine functions (listed in the order):

@@

@@

Given the SineModel5 parameterized motion model (see previous section) and the allowable ranges for its five
parameters (\tabref{parameters}), the task is discovering
values for the five parameters that result in fast 
gaits.

@@

We evaluated the following six different
learning algorithms for the parameterized motion models:

@@

@@

\emph{Random}: This method randomly generates parameter vectors in the allowable range for every trial.  This strategy serves as as baseline for comparison.

\emph{Uniform random hill climbing}: This method always starts
  with the current best gait, and then chooses the next $\vec{\theta}$ by
  randomly choosing one parameter to adjust and replacing it with a
  new value chosen with uniform probability in the allowable range for
  that parameter. This new point is evaluated, and if it results in a
  longer distance walked than the previous best gait, it is saved as
  the new best gait. The process is then repeated.

@@

\emph{N-dimensional policy gradient ascent}: We implemented Kohl
  and Stone's \citep{kohl} method for local gradient ascent for gait
  learning with noisy fitness evaluations. This strategy explicitly
  estimates the gradient of the objective function. It does this by
  first generating $n$ parameter vectors near the initial vector by
  perturbing each dimension of each vector randomly by either
  $-\epsilon$, $0$, or $\epsilon$. Then each vector is run on the
  robot, and for each dimension we segment the results into three
  groups: $-\epsilon$, $0$, and $\epsilon$.  The gradient along this
  dimension is then estimated as the average score for the $\epsilon$
  group minus the average score for the $-\epsilon$ group. Finally,
  the method creates the next $\vec{\theta}$ by changing all
  parameters by a fixed-size step in the direction of the gradient.
  For this study we used values of $\epsilon$ equal to 5\% of the
  allowable range in each dimension (ranges listed in
  \tabref{parameters}), and a step size scaled such that if all dimensions
  were in the range [0,~1], the norm of the step size would be 0.1.

@@

Three runs were performed per learning method. To most directly compare learning methods, we
evaluated the different methods by starting each of their three runs, respectively, with the same three randomly-chosen initial parameter vectors~($\vec{\theta}_A$,
$\vec{\theta}_B$, and $\vec{\theta}_C$). Runs continued until the performance plateaued, which we defined as when there was no improvement during the last third of
a run.

### 2011-04-14 paper_ecal/platformDetails.tex
More edits, 75% done :)
Cleaned a few things up
more forging

\subsection{Platform details}
\seclabel{platformDetails}

@@

The quadruped robot used in this study was assembled from parts
printed on the Objet Connex 500 3-D Printing System. It weighs 1.88 kg
with the on-board computer and measures approximately 38 centimeters
from leg to opposite leg in the crouch position depicted in
\figref{robot_whitebg.jpg}. The robot is actuated by 9 AX-12+
Dynamixel servos: one inner joint and one outer joint servo in each of
the four legs, and one servo at the center ``hip'' joint.  This final
unique servo allows the two halves of the robot to rotate with respect
to each other. \figref{topdown2.jpg} shows this unique motion, as well
as the positions and numerical designations of all nine servos.  Each
servo could be commanded to a position in the range [0,~1023],
corresponding to a physical range [-120$^{\circ}$,~+120$^{\circ}$].

All of the computation for gait learning, fitness evaluation, and
robot control was performed on the compact, on-board CompuLab Fit-PC2,
running Ubuntu Linux 10.10. 
The most computationally expensive portion of code was HyperNEAT, but even
this took less than one second to run on the on-board computer. Thus,
we chose not to offload any computation.
All gait generation, learning, and fitness
evaluation code, except HyperNEAT, was written in Python and is
available on our website \citep{quadratotWeb}.  HyperNEAT is written in
C++.  We controlled the servos with the \code{pydynamixel} library
\citep{pydynamixel}, sending commanded positions at 40Hz.  The robot
connected to a wireless network on boot, which enabled us to control
it via SSH.

To run a gait on the robot, one needed only to provide a \emph{gait
  function} --- any Python function that accepts a single input, time
starting at 0, and outputs a list of nine commanded positions, one for
each servo.  To safeguard against limb collision with the robot body,
the robot control code cropped the commands to a safe range.  This
range was [150,~770] for the inner leg servos, [30,~680] for the outer
leg servos, and [392,~623] for the center hip servo.

>>>>

\subsection{Platform Details}
\seclabel{platformDetails}

@@

The quadruped robot in this study was assembled from off-the-shelf components and parts
printed on the Objet Connex 500 3-D Printing System. It weighs 1.88 kg
with the on-board computer and measures approximately 38 centimeters
from leg to opposite leg in the crouch position depicted in
\figref{robot_whitebg.jpg}. The robot is actuated by 9 AX-12+
Dynamixel servos: one inner joint and one outer joint servo in each of
the four legs, and one servo at the center ``hip'' joint.  This final
unique servo allows the two halves of the robot to rotate with respect
to each other. \figref{topdown2.jpg} shows this unique motion, as well
as the positions and numerical designations of all nine servos.  Each
servo could be commanded to a position in the range [0,~1023],
corresponding to a physical range [-120$^{\circ}$,~+120$^{\circ}$].
The computer and servos can be powered by two on-board batteries, but
for the tests presented in this paper power was provided by a tethered
cable.

All of the computation for gait learning, fitness evaluation, and
robot control was performed on the compact, on-board CompuLab Fit-PC2,
running Ubuntu Linux 10.10. 
The most computationally expensive portion of code was HyperNEAT, but even
this took less than one second to run on the on-board computer. Thus,
we chose not to offload any computation.
All gait generation, learning, and fitness
evaluation code, except HyperNEAT, was written in Python and is
available on our website \citep{quadratotWeb}.  HyperNEAT is written in
C++.  We controlled the servos with the Pydynamixel library
\citep{pydynamixel}, sending commanded positions at 40Hz.  The robot
connected to a wireless network on boot, which enabled us to control
it via SSH.

Robot gaits are executed via a Python \emph{gait function} that takes time (starting at 0) as a single input and outputs a list of nine commanded positions (one for each servo).  To safeguard against limb collision with the robot body,
the robot control code cropped the commands to a safe range.  This
range was [150,~770] for the inner leg servos, [30,~680] for the outer
leg servos, and [392,~623] for the center hip servo.

### 2011-04-14 paper_ecal/problemDefinition.tex
more forging

The gait learning problem aims to find a \emph{gait} that maximizes
some performance metric.

Mathematically, we define a gait as a function that specifies
a vector of commanded motor positions for a robot over time.
%Using this definition, we require one function per motor.
We can write gaits without feedback --- also called open-loop or
feedforward gaits --- as

@@

It follows that feedforward gaits are deterministic, producing the
same command pattern each time they are run.  While the commanded
positions will be the same from trial to trial, the actual robot
motion and measured fitness will vary due to the noisiness of trials
in the real world.

@@

The ultimate goal was to design gaits that were as fast as possible,
so to optimize for speed, our performance metric was displacement over
a period of motion lasting approximately 12 seconds. Details of
exactly how this displacement was measured are given below.

>>>>

The gait learning problem aims to find a \emph{gait} that maximizes
some performance metric. Mathematically, we define a gait as a function that specifies
a vector of commanded motor positions for a robot over time.
%Using this definition, we require one function per motor.
We can write gaits without feedback --- also called open-loop gaits --- as

@@

It follows that open-loop gaits are deterministic, producing the
same command pattern each time they are run.  While the commanded
positions will be the same from trial to trial, the actual robot
motion and measured fitness will vary due to the noisiness of trials
in the real world.

@@

The ultimate goal was to design gaits that were as fast as possible.
Our performance metric was thus displacement over
the evaluation period of 12 seconds. Details of
how this displacement was measured are given below.
